Assignment 1. Activity Monitoring Data
======================================

###Loading and preprocessing the data

1. Firstly, let's read csv file from the web:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
#Loading library
library(plyr)
library(ggplot2)
library(gridExtra)
```

```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
#Creating temporary file in memory
temp <- tempfile()
#Just to use download.file in R Markdown
setInternet2(use = TRUE)
#Download zip file in temp
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp)
#Unzip temp file and read "activity.csv" from it to activity variable
activity <- read.csv(unz(temp, "activity.csv"))
#Delete temp file
unlink(temp)
```

2. Explore size of data.frame:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
dim(activity)
```

3. Explore type of variables in **activity** data.frame:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
str(activity)
```
So, **steps** variable is correct type. It is better to change type of **date** to some date type. **interval** variable represents minutes interval. It is better to transform this column to two columns, one of which will represent start hour, second - start minute.

```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
#Tranforming date variable to Date type
activity$date <- as.Date(activity$date)

#Adding two variables based on interval variable
activity$interval <- as.character(activity$interval)
activity$interval <- sapply(activity$interval, FUN = function(element) return(paste(as.character(c(rep(0, (4-nchar(element))), element)), collapse = "")))

activity$start_hour <- as.integer(sapply(activity$interval, FUN = function (element) return (substr(element, 1, 2))))
activity$start_minute <- as.integer(sapply(activity$interval, FUN = function (element) return (substr(element, 3, 4))))
```

Now, data looks like:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
head(activity)
```

And type of variables:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
str(activity)
```

***

###What is mean total number of steps taken per day?

1. Total number of steps per day:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
total_steps <- ddply(activity, ~date, summarise, Total_steps=sum(steps, na.rm = TRUE))
total_steps
```

2. Histogram of total number of steps:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
rng <- range(total_steps$Total_steps)
p1 <- qplot(Total_steps, 
      data=total_steps,
      geom="histogram", 
      breaks = seq(rng[1], rng[2], by =1700),
      main = "Histogram of number of\n steps per day",
      xlab = "Steps per day",
      ylab = "Days"
      )
p1
```

3. Mean and median of the total number of steps taken per day:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
#Mean number of total number of steps per day
mean(total_steps$Total_steps, na.rm = TRUE)
#Median number of total number of steps per day
median(total_steps$Total_steps, na.rm = TRUE)
```

***
###What is the average daily activity pattern?

1. Average number of steps per 5 minute interval:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
activity$minute_from_00 <- activity$start_hour * 60 + activity$start_minute
avr_steps_5min <- ddply(activity, ~minute_from_00, summarise, Avr_steps=mean(steps, na.rm = TRUE))
head(avr_steps_5min)
```

2. Plotting average values per 5 minutes:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
p2 <- qplot(minute_from_00,
      Avr_steps,
      data=avr_steps_5min,
      geom=c("point", "path"),
      main = "Average of number of\n steps per 5 minute interval",
      xlab = "Minutes from 00:00",
      ylab = "Average number of steps"
      )
p2
```

3. 5 minutes interval with max number of average steps:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
avr_steps_5min[which.max(avr_steps_5min$Avr_steps),]
```
So, max number of steps at average is taken place from 08:35 to 08:40.

***

###Imputing missing values

1. Total number of missing values in the dataset:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
#Number of NAs in activity data.frame
apply(activity[,1:3], 2, FUN = function(column) return(sum(is.na(column))))
```
So, there are 2304 NAs in **steps** column and no NAs in **date** and **interval** columns.

2. Replacing NAs in **steps** column with average number of steps per 5 minute interval:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
activity_noNA <- activity

for (i in 1:nrow(activity_noNA)) {
        if (is.na(activity_noNA$steps[i])) {
                activity_noNA$steps[i] <- avr_steps_5min$Avr_steps[avr_steps_5min$minute_from_00 == activity_noNA$minute_from_00[i]]
        }
} 

#Number of NAs in activity_noNA data.frame
apply(activity_noNA[,1:3], 2, FUN = function(column) return(sum(is.na(column))))
```
There are no NAs left.

Head of the data now looks like:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
head(activity_noNA)
```

3. Total number of steps taken each day, calculated based on data with imputed NA:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
total_steps_noNA <- ddply(activity_noNA, ~date, summarise, Total_steps=sum(steps, na.rm = TRUE))
head(total_steps_noNA)
```

Two histogram of total number of steps without and with imputed NAs:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
rng <- range(total_steps_noNA$Total_steps)
p3 <- qplot(Total_steps, 
      data=total_steps_noNA,
      geom="histogram", 
      breaks = seq(rng[1], rng[2], by =1700),
      main = "Histogram of number of\n steps per day with imputed NA",
      xlab = "Steps per day",
      ylab = "Days")
grid.arrange(p1, p3, ncol=2)
```

4. Mean and median of the total number of steps taken per day without and with imputed NAs:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
#Mean number of total number of steps per day
mean(total_steps$Total_steps, na.rm = TRUE)
mean(total_steps_noNA$Total_steps)
#Median number of total number of steps per day
median(total_steps$Total_steps, na.rm = TRUE)
median(total_steps_noNA$Total_steps)
```

Imputing has next impact:  
1. Histogram changed its view a little bit. Now, there are no days with no steps, but more days with ~11000 steps. It as because we changed total number of steps for 8 days;  
2. Mean and Medium values of steps became the same -> distribution became more normal.

***

###Are there differences in activity patterns between weekdays and weekends?

1. Creating a new factor variable with weekday and more for identifying weekend:
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
Sys.setlocale("LC_TIME", "English")
```

```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
activity_noNA$weekday <- factor(weekdays(activity_noNA$date))
activity_noNA$weekend <- factor(ifelse((activity_noNA$weekday == "Saturday" | activity_noNA$weekday == "Sunday"), "weekend", "weekday"))
```


2. Creating data for weekdays and weekends:

```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
avr_steps_5min_weekends <- ddply(activity_noNA, .(minute_from_00, weekend), summarise, Avr_steps=mean(steps))
head(avr_steps_5min_weekends)
```

Visualizing with one plot, but two series (it is the same as using panel plots):
```{r echo=TRUE, warning=FALSE, info = FALSE, comment=""}
qplot(minute_from_00,
      Avr_steps,
      color = weekend,
      data=avr_steps_5min_weekends,
      geom=c("point", "path"),
      main = "Average of number of\n steps per 5 minute interval",
      xlab = "Minutes from 00:00",
      ylab = "Average number of steps"
      )
```

Very interesting pattern could be observed: during weekdays average number of steps is higher for morning hours, where for weekends - day hours. For the evening hours - patterns are very similar, but ~ from 19:00 till 21:00 average number of steps is higher for weekends.